\section{Introduction}
\label{sec:intro}

Synthetic media has progressed from single-modality manipulation to coordinated audio-visual (AV) forgery, in which a face, a voice, and their temporal alignment are all altered. Detectors have followed, moving from per-stream artifact recognition to cross-modal consistency reasoning, and they increasingly consume \emph{frozen} self-supervised representations such as DINOv2~\cite{oquab2024dinov2} and Whisper~\cite{radford2023whisper} rather than raw pixels and waveforms. These detectors are accurate on clean data, but their robustness is almost always established empirically, against a fixed catalogue of attacks and degradations. In high-stakes forensic settings this is not enough: a defensible verdict needs a \emph{per-sample guarantee} that a small perturbation cannot flip the decision.

Randomized smoothing~\cite{cohen2019certified} provides exactly such a guarantee for image classifiers, certifying that the prediction is constant within an $\ell_2$ ball. Porting it to AV deepfake detection is non-obvious: clips are long, audio is continuous, and the detector never sees raw input --- it consumes a cached, frozen feature vector. CertAV therefore certifies the detector in this joint feature space. The guarantee is deliberately scoped: it is exact for any adversary that perturbs the representation consumed by the detector, but it is not an end-to-end pixel/waveform certificate. We argue this is the right frontier, not a compromise: as detectors are increasingly assembled from frozen foundation encoders, the representation --- not the raw signal --- is the interface an attacker and a defender actually share, and end-to-end certification of such encoders remains computationally out of reach. We nonetheless bridge the two spaces empirically (Sec.~\ref{sec:method},~\ref{sec:exp}) by composing the feature certificate with a measured feature-to-input Lipschitz constant.

Our starting empirical observation is striking: feature-space smoothing produces \emph{large} certified radii with \emph{negligible} abstention, and this is largely independent of whether noise augmentation is used during training. This is unusual --- in pixel space, large radii usually require heavy noise training and incur substantial clean-accuracy loss. We trace the effect to geometry. The frozen multimodal representation occupies a low-dimensional subspace of its ambient feature space ($\approx\!80$ of $768$ dimensions carry $90\%$ of the variance). If the detector's decision depends only on this subspace, then isotropic Gaussian noise can move a sample far in the many irrelevant directions without ever crossing the boundary. We formalize this as a \emph{Certifiability Scaling} theorem and show that the gap between an adversary's ambient budget and its on-manifold effect scales as $\sqrt{D/d}$. This single mechanism (i) explains the large radii, (ii) predicts which encoders certify best, (iii) explains why a no-noise model certifies almost as well as a noise-trained one, and (iv) motivates a smoothing distribution \emph{aligned} to the manifold.

The narrative of this paper is therefore coherent rather than incremental: an empirical observation about geometry leads to a theorem, the theorem is validated across encoders, and the same geometry yields a stronger, complementary guarantee. Concretely, we make five contributions.

\begin{enumerate}
\itemsep0.15em
\item \textbf{CertAV pipeline.} A feature-space randomized-smoothing certification pipeline for AV deepfake detection over frozen DINOv2 and Whisper encoders, under a joint $\ell_2$ threat model (Sec.~\ref{sec:method}).
\item \textbf{Certifiability Scaling theorem.} A proof that under subspace invariance the certified radius depends only on the $d$-dimensional on-manifold noise, with amplification $\sqrt{D/d}$ and a corollary that gives an encoder-selection rule (Sec.~\ref{sec:theory}).
\item \textbf{Encoder-scaling study.} An empirical validation across five visual/audio encoder families confirming that lower $d/D$ yields larger certified radii (Sec.~\ref{sec:exp}).
\item \textbf{Manifold-aware anisotropic smoothing.} A PCA-aligned anisotropic smoothing scheme that, at equalized noise budget, delivers a $3.4\times$ larger on-manifold certified radius with zero abstention (Sec.~\ref{sec:exp}).
\item \textbf{Robust conformal prediction.} A distributional coverage guarantee for \emph{all} samples --- including those CertAV abstains on --- using the smoothed classifier's confidence bounds as nonconformity scores (Sec.~\ref{sec:exp}).
\end{enumerate}
