\section{Related Work}
\label{sec:related}

\paragraph{Audio-visual deepfake detection.}
Cross-modal detectors exploit inconsistencies that are invisible within a single stream. Early work used affective cues~\cite{mittal2020emotions} and audio-visual dissonance~\cite{chugh2020not}; later methods turned to self-supervised anomaly detection~\cite{feng2023self}, joint AV learning~\cite{yang2023avoid}, and local temporal inconsistency modeling~\cite{astrid2025local}. Benchmarks such as FakeAVCeleb~\cite{khalid2021fakeavceleb} and LAV-DF~\cite{cai2022lavdf} drive this progress with multimodal manipulations of the visual and/or audio stream. These methods steadily improve clean accuracy and forgery localization, but their robustness is reported empirically; none provide a per-sample certificate. CertAV is complementary: it certifies a detector built on the same frozen encoders.

\paragraph{Certified robustness via randomized smoothing.}
Randomized smoothing~\cite{cohen2019certified} certifies an $\ell_2$ ball by lower-bounding the probability that a Gaussian-perturbed base classifier returns the top class; related lines connect it to differential privacy~\cite{lecuyer2019certified} and to adversarially trained smoothed classifiers~\cite{salman2019provably}. Yang et al.~\cite{yang2020randomized} generalize smoothing to broad noise families and show that non-isotropic distributions certify \emph{ellipsoidal} regions, and ANCER~\cite{eiras2022ancer} optimizes a per-sample anisotropic covariance to maximize certified \emph{volume}. MMCert~\cite{wang2024mmcert} is the first certified defense for multimodal models, using randomized ablation of raw inputs to bound robustness to bounded per-modality corruption. CertAV differs on three axes: (i) it certifies the frozen \emph{feature} space consumed by the detector rather than raw input (the natural abstraction for frozen-encoder pipelines), (ii) it provides a theoretical account of \emph{why} this space certifies unusually well, and (iii) its anisotropic distribution is a single global PCA-aligned covariance justified by that account, rather than a per-sample volume optimization. MMCert and CertAV are complementary: the former covers the encoder-input boundary, the latter the detector-input boundary.

\paragraph{Conformal prediction.}
Conformal prediction~\cite{angelopoulos2023conformal,romano2020classification} converts any score into prediction \emph{sets} with a finite-sample marginal coverage guarantee, and Gendler et al.~\cite{gendler2022adversarially} extend this to adversarially robust coverage by inflating nonconformity scores under bounded perturbation. We instantiate this framework for AV forensics, using the smoothed classifier's Clopper-Pearson lower bounds as scores --- a natural bridge from smoothing certificates to conformal coverage that, to our knowledge, has not been explored for multimodal deepfake detection. The benefit is operational: CertAV abstains when the certified probability falls below $0.5$, whereas conformal prediction still returns a (possibly two-element) set with guaranteed coverage for those samples.
